---
title: "RMS Titanic"
author: "Isa de Luis"
date: 2025-30-1
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class
-   Sex
-   Age
-   Survived
-   n
    -   What is `n`? The other four variables are self-explanatory, but `n` is not. Looking at the description of the dataset, it does not include `n`, so I assume it was created when turning the dataset into a tibble?

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

```{r}
?Titanic
```

**Observations**:

-   Write your observations here
    -   The wikipedia article states there were 2,224 passengers and crew aboard the titanic
    -   The dataset has 2,201 passengers -\> missing 23 passengers
    -   According to details of the dataset, the data was originally collected by the British Board of Trade in their investigation of the sinking
        -   Says that there is not a complete agreement among sources as to the exact number on board, rescued, or lost
-   Are there any differences?
    -   Yes, the dataset has 23 less passengers than were actually on the Titanic

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`

df_titanic |>
  filter(Survived == "Yes") |>
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge") +
  xlab("Class") +
  ylab("Number of People") +
  ggtitle("Class and Sex of Survivors of the Titanic Crash")
    
```

**Observations**:

-   The number of women who survived **always** outnumbered the number of men who survived for each class, except for the crew, however, the crew was most likely overwhelmingly male, explaining the discrepancy
-   In third class, the number of men and women who survived are almost equal

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop |>
  filter(Survived == "Yes") |>
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge", color = "Black") +
  xlab("Class") +
  ylab("Proportion of People") +
  ggtitle("Proportion of Survivors by Class and Sex")
```

**Observations**:

-   Write your observations here.
    -   Super high survival rate in first and second class
    -   In third class, men have a low survival rate compared to women
    -   Overall third class has a much lower survival rate compared to 1st and 2nd class
    -   Female crew had a much higher survival rate than male crew
-   Is there anything *fishy* going on in your plot?
    -   First and second class proportions are exactly one – something is odd with that, because it suggests every person in these groups survived
        -   Is that plausible?

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop |>
  filter(Survived == "Yes") |>
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap("Age") + 
  xlab("Class") +
  ylab("Proportion of People") +
  ggtitle("Proportion of Survivors by Class, Sex, and Age")
```

**Observations**:

-   All 1st and 2nd class children survived
-   3rd class girls were more likely to survive than 3rd class boys
-   Almost all 1st class women survived
-   In all class, more women than men survived
-   2nd class men were least likely to survive
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   The fishiness is because the bars are stacked, and without an outline you can't distinguish the stacks

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
