---
title: "Gapminder"
author: "Isa de Luis"
date: 2025-12-02
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
?gapminder
summary(gapminder)
head(gapminder)
```

**Observations**:

-   The dataset has data from 5 continents (North America and South America are combined)

-   The average life expectancy is 59.47

-   The average GDP is 7215.3

-   The dataset has 6 variables:

    -   Country

    -   Continent

    -   Year

    -   Life Expectancy

    -   Population

    -   GDP Per Capita

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  gapminder |>
  pull(year) |> 
  max()

year_min <- 
  gapminder |>
  pull(year) |>
  min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder |>
  filter(year == year_min) |>
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot()
```

**Observations**:

-   Almost all of the continents have some outliers, except Oceania.
-   Asia has the widest range of GDP per Capita
-   Oceania has the smallest range of GDP per Capita

**Difficulties & Approaches**:

-   At first I tried a bar plot, but I found that I couldn't see the individual countries with a bar plot, which made question 3 difficult, since I couldn't identify outliers
-   Then, I tried a boxplot, which was helpful for finding outliers, but the boxes were very condensed because of the one outlier in Asia that's over 90000 GDP per capita

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2

gapminder |>
  group_by(continent) |>
  filter(year == year_min) |>
  mutate(
    outlier_range = (quantile(gdpPercap, probs = .75) + (1.5 * IQR(gdpPercap)))) |>
  filter(gdpPercap > outlier_range)
  
```

**Observations**:

-   Identify the outlier countries from q2
    -   Angola
    -   Bahrain
    -   Canada
    -   Gabon
    -   Kuwait
    -   South Africa
    -   Switzerland
    -   US
    -   Venezuela

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%
  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder_filtered <- 
  gapminder |>
  group_by(continent) |>
  filter(year == year_min | year == year_max) 

gapminder_filtered |>
  ggplot(aes(x = continent, y = gdpPercap)) + 
  geom_boxplot() + 
  geom_point(
    data = gapminder_filtered |> filter(country %in% c("Angola",
                                                       "Bahrain",
                                                       "Canada",
                                                       "Gabon",
                                                       "Kuwait",
                                                       "South Africa",
                                                       "Switzerland",
                                                       "United States",
                                                       "Venezula")),
    mapping = aes(color = country),
    size = 2
  ) + 
  facet_wrap("year") 
```

**Observations**:

-   Across most continents, median GDP per capita has increased
-   In 2007, the spread within continents increased
-   Africa had the lowest median GDP per capita in both 1952 and 2007
-   In 1952, GDP per capita was quite low, though Kuwait was a notable outlier
    -   By 2007, Kuwait was still an outlier, but not as large as an outlier
-   Oceania continuously has the highest GDP per capita
-   Europe's GDP rose the most between 1952 and 2007

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## Comparing populations in min and max year
gapminder |>
  filter(year == year_min | year == year_max) |>
  group_by(continent) |>
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  facet_wrap("year") 
```

-   Median life expectancy rose in every continent
-   The minimum life expectancy rose in every country
-   Is the life expectancy rising correlated to GDP?
-   How does the change in life expectancy relate to change in GDP over time?

```{r q5-task2}
## TASK: Your second graph
# find country with greatest difference in life expectancy between 1952 and 2007
gapminder_1952 <- gapminder |>
  filter(year == year_min) |>
  mutate(lifeExp_1952 = lifeExp) |>
  select(country, lifeExp_1952)

gapminder_2007 <- gapminder |>
  filter(year == year_max) |>
  mutate(lifeExp_2007 = lifeExp) |>
  select(country, lifeExp_2007)

# Merge df
gapminder_merged <-
  merge(gapminder_1952, gapminder_2007) |>
  mutate(diff = lifeExp_2007 - lifeExp_1952) |>
  select(-c(lifeExp_2007, lifeExp_1952))

gapminder_merged
country_names <- 
  gapminder_merged |>
  arrange(column = desc(diff)) |>
  pull(country)

gapminder |>
  filter(country == country_names[1]) |>
  ggplot(aes(x = year, y = lifeExp, color = gdpPercap)) +
  geom_point()
```

-   Higher life expectancy is directly correlated to higher GDP
-   Higher life expectancy is not linearly related to higher GDP
-   Does this hold true for other countries?
-   How is population influenced by GDP? Life Expectancy?

```{r q5-task3}
## TASK: Your third graph
# look at correlation between difference in gdp and difference in life expectancy in each country
# expect < diff in gdp to lead to < diff in life expectancy

gapminder_1952 <- gapminder |>
  filter(year == year_min) |>
  mutate(lifeExp_1952 = lifeExp,
         gdp_1952 = gdpPercap) |>
  select(country, continent, lifeExp_1952, gdp_1952)



gapminder_2007 <- gapminder |>
  filter(year == year_max) |>
  mutate(lifeExp_2007 = lifeExp,
         gdp_2007 = gdpPercap) |>
  select(country, continent, lifeExp_2007, gdp_2007)

gapminder_merged <-
  merge(gapminder_1952, gapminder_2007) |>
  mutate(diff_lifeExp = lifeExp_2007 - lifeExp_1952,
         diff_gdp = gdp_2007 - gdp_1952) |>
  select(-c(lifeExp_2007, lifeExp_1952, gdp_2007, gdp_1952))

# plots for each continent
gapminder_merged |>
  filter(continent == "Africa") |>
  ggplot(aes(x = country, y = diff_lifeExp, color = diff_gdp)) +
  geom_point(size = 2) +
  scale_colour_gradient2(low = "red", mid = "green", high = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

gapminder_merged |>
  filter(continent == "Asia") |>
  ggplot(aes(x = country, y = diff_lifeExp, color = diff_gdp)) +
  geom_point(size = 2) +
  scale_colour_gradient2(low = "red", mid = "green", high = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

gapminder_merged |>
  filter(continent == "Americas") |>
  ggplot(aes(x = country, y = diff_lifeExp, color = diff_gdp)) +
  geom_point(size = 2) +
  scale_colour_gradient2(low = "red", mid = "green", high = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

gapminder_merged |>
  filter(continent == "Europe") |>
  ggplot(aes(x = country, y = diff_lifeExp, color = diff_gdp)) +
  geom_point(size = 2) +
  scale_colour_gradient2(low = "red", mid = "green", high = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
gapminder_merged |>
  filter(continent == "Oceania") |>
  ggplot(aes(x = country, y = diff_lifeExp, color = diff_gdp)) +
  geom_point(size = 2) +
  scale_colour_gradient2(low = "red", mid = "green", high = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

-   Most African countries show a positive increase in life expectancy over the observed period
-   Life expectancy improvements vary widely, with some countries seeing increases above 30 years, while others have smaller improvements under 10 years
-   Most Asian countries experienced increases in life expectancy between \~15 to 35 years
-   Most countries in the Americas saw life expectancy increases between 10 to 30 years
-   European countries generally saw smaller increases in life expectancy compared to other continents, mostly between 10 to 20 years
-   Some Eastern European countries (like Albania) show higher increases than Western European ones
-   In general, countries with positive GDP growth also saw increases in life expectancy, but the relationship is not strictly linear
-   Outliers like Lebanon show that life expectancy can increase despite declining GDP
