---
title: "US Income"
author: "Isa de Luis"
date: 2025-04-06
output:
  github_document:
    toc: true
prerequisites:
  - e-stat09-bootstrap
---

*Purpose*: We've been learning how to quantify uncertainty in estimates through the exercises; now its time to put those skills to use studying real data. In this challenge we'll use concepts like confidence intervals to help us make sense of census data.

*Reading*: - [Using ACS Estimates and Margin of Error](https://www.census.gov/data/academy/webinars/2020/calculating-margins-of-error-acs.html) (Optional, see the PDF on the page) - [Patterns and Causes of Uncertainty in the American Community Survey](https://www.sciencedirect.com/science/article/pii/S0143622813002518?casa_token=VddzQ1-spHMAAAAA:FTq92LXgiPVloJUVjnHs8Ma1HwvPigisAYtzfqaGbbRRwoknNq56Y2IzszmGgIGH4JAPzQN0) (Optional, particularly the *Uncertainty in surveys* section under the Introduction.)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

# Setup

<!-- ----------------------------------------------------------------------- -->

```{r setup}
library(tidyverse)
```

### **q1** Load the population data from c06; simply replace `filename_pop` below.

```{r q1-task}
## TODO: Give the filename for your copy of Table B01003
filename_pop <- "data/ACSDT5Y2018.B01003_2025-03-10T161941/county-data.csv"

## NOTE: No need to edit
df_pop <-
  read_csv(
    filename_pop,
    skip = 1,
  ) %>% 
  rename(
    population_estimate = `Estimate!!Total`
  )

df_pop
```

You might wonder why the `Margin of Error` in the population estimates is listed as `*****`. From the [documentation (PDF link)](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&cad=rja&uact=8&ved=2ahUKEwj81Omy16TrAhXsguAKHTzKDQEQFjABegQIBxAB&url=https%3A%2F%2Fwww2.census.gov%2Fprograms-surveys%2Facs%2Ftech_docs%2Faccuracy%2FMultiyearACSAccuracyofData2018.pdf%3F&usg=AOvVaw2TOrVuBDlkDI2gde6ugce_) for the ACS:

> If the margin of error is displayed as ‘\*\*\*\*\*’ (five asterisks), the estimate has been controlled to be equal to a fixed value and so it has no sampling error. A standard error of zero should be used for these controlled estimates when completing calculations, such as those in the following section.

This means that for cases listed as `*****` the US Census Bureau recommends treating the margin of error (and thus standard error) as zero.

### **q2** Obtain median income data from the Census Bureau:

-   `Filter > Topics > Income and Poverty > Income and Poverty`
-   `Filter > Geography > County > All counties in United States`
-   Look for `Median Income in the Past 12 Months` (Table S1903)
-   Download the 2018 5-year ACS estimates; save to your `data` folder and add the filename below.

```{r q2-task}
## TODO: Give the filename for your copy of Table S1903
filename_income <- "data/ACSST5Y2018.S1903_2025-04-06T190013/data.csv"

## NOTE: No need to edit
df_income <-
  read_csv(filename_income, skip = 1)
df_income
```

Use the following test to check that you downloaded the correct file:

```{r q2-tests}
## NOTE: No need to edit, use to check you got the right file.
assertthat::assert_that(
  df_income %>%
    filter(Geography == "0500000US01001") %>%
    pull(`Estimate!!Percent Distribution!!FAMILY INCOME BY FAMILY SIZE!!2-person families`)
  == 45.6
)

print("Well done!")
```

This dataset is in desperate need of some *tidying*. To simplify the task, we'll start by considering the `\\d-person families` columns first.

### **q3** Tidy the `df_income` dataset by completing the code below. Pivot and rename the columns to arrive at the column names `id, geographic_area_name, category, income_estimate, income_moe`.

*Hint*: You can do this in a single pivot using the `".value"` argument and a `names_pattern` using capture groups `"()"`. Remember that you can use an OR operator `|` in a regex to allow for multiple possibilities in a capture group, for example `"(Estimate|Margin of Error)"`.

```{r q3-task}

df_q3 <-
  df_income %>%
  select(
    Geography,
    contains("Geographic"),
    # This will select only the numeric d-person family columns;
    # it will ignore the annotation columns
    contains("median") & matches("\\d-person families") & !contains("Annotation of")
  ) %>%
  mutate(across(contains("median"), as.numeric)) %>%
## TODO: Pivot the data, rename the columns
  pivot_longer(
    cols = -c(Geography, `Geographic Area Name`),
    names_to = c("category", ".value"),
    names_pattern = "(Estimate|Margin of Error)!!Median income \\(dollars\\)(?: MOE)?!!FAMILY INCOME BY FAMILY SIZE!!(.*)"
  ) |> pivot_longer(
    cols = `2-person families`:`6-person families`,
    names_to = "fam_size",
    values_to = "val"
  ) %>%
  # 2) Pivot "Estimate"/"Margin of Error" from rows to columns
  pivot_wider(
    names_from = category,     # "Estimate" or "Margin of Error"
    values_from = val,         # numeric values
    names_prefix = "income_"
  ) %>%
  # 3) Rename to final tidy names
  rename(
    Geography = Geography,
    geographic_area_name = `Geographic Area Name`,
    category = fam_size,
    income_estimate = income_Estimate,
    income_moe = `income_Margin of Error`
  ) %>%
  glimpse()

```

Use the following tests to check your work:

```{r q3-tests}
## NOTE: No need to edit
assertthat::assert_that(setequal(
  names(df_q3),
  c("Geography", "geographic_area_name", "category", "income_estimate", "income_moe")
))

assertthat::assert_that(
  df_q3 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_moe)
  == 6663
)

print("Nice!")
```

The data gives finite values for the Margin of Error, which is closely related to the Standard Error. The Census Bureau documentation gives the following relationship between Margin of Error and Standard Error:

$$\text{MOE} = 1.645 \times \text{SE}.$$

### **q4** Convert the margin of error to standard error. Additionally, compute a 99% confidence interval on income, and normalize the standard error to `income_CV = income_SE / income_estimate`. Provide these columns with the names `income_SE, income_lo, income_hi, income_CV`.

```{r q4-task}

df_q4 <- df_q3 %>%
  mutate(
    income_SE = income_moe / 1.645,
    income_lo = income_estimate - (2.576 * income_SE),
    income_hi = income_estimate + (2.576 * income_SE),
    income_CV = income_SE / income_estimate
  )

df_q4
```

Use the following tests to check your work:

```{r q4-tests}
## NOTE: No need to edit
assertthat::assert_that(setequal(
  names(df_q4),
  c("Geography", "geographic_area_name", "category", "income_estimate", "income_moe",
    "income_SE", "income_lo", "income_hi", "income_CV")
))

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_SE) - 4050.456
  ) / 4050.456 < 1e-3
)

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_lo) - 54513.72
  ) / 54513.72 < 1e-3
)

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_hi) - 75380.28
  ) / 75380.28 < 1e-3
)

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_CV) - 0.06236556
  ) / 0.06236556 < 1e-3
)

print("Nice!")
```

One last wrangling step: We need to join the two datasets so we can compare population with income.

### **q5** Join `df_q4` and `df_pop`.

```{r q5-task}
## TODO: Join df_q4 and df_pop by the appropriate column

df_data <-
  df_q4 %>%
  left_join(df_pop, by = "Geography")
```

# Analysis

<!-- ----------------------------------------------------------------------- -->

We now have both estimates and confidence intervals for `\\d-person families`. Now we can compare cases with quantified uncertainties: Let's practice!

### **q6** Study the following graph, making sure to note what you can *and can't* conclude based on the estimates and confidence intervals. Document your observations below and answer the questions.

```{r q6-task}
## NOTE: No need to edit; run and inspect
wid <- 0.5

df_data %>%
  filter(str_detect(geographic_area_name, "Massachusetts")) %>%
  mutate(
    county = str_remove(geographic_area_name, " County,.*$"),
    county = fct_reorder(county, income_estimate)
  ) %>%

  ggplot(aes(county, income_estimate, color = category)) +
  geom_errorbar(
    aes(ymin = income_lo, ymax = income_hi),
    position = position_dodge(width = wid)
  ) +
  geom_point(position = position_dodge(width = wid)) +

  coord_flip() +
  labs(
    x = "County",
    y = "Median Household Income"
  )
```

**Observations**:

-   Document your observations here.
    -   Larger family sizes often show higher median income than smaller household sizes
    -   Some counties have noticeably wider error bars, suggesting more variability or smaller sample sizes.
-   Can you confidently distinguish between household incomes in Suffolk county? Why or why not?
    -   No! The confidence intervals for Suffolk County often overlap with those of other counties (and sometimes across different family sizes within Suffolk itself). That overlap means there’s no statistically significant difference at the chosen confidence level for many of these comparisons.
-   Which counties have the widest confidence intervals?
    -   Dukes and Nantucket

In the next task you'll investigate the relationship between population and uncertainty.

### **q7** Plot the standard error against population for all counties. Create a visual that effectively highlights the trends in the data. Answer the questions under *observations* below.

*Hint*: Remember that standard error is a function of *both* variability (e.g. variance) and sample size.

```{r q7-task}
df_data %>%
  mutate(pop_bin = cut_number(population_estimate, n = 5)) %>%
  ggplot(aes(x = pop_bin, y = income_SE, fill = category)) +
  geom_boxplot() +
  labs(
    title = "Standard Error Across Population Bins",
    x = "Population Bins (Quintiles)",
    y = "Income Standard Error"
  ) +
  theme_minimal()
```

**Observations**:

-   What *overall* trend do you see between `SE` and population? Why might this trend exist?
    -   The overall trend isi that as population size increases, the standard error decreases. This trend might exist because the standard error (SE) is partly determined by sample size—larger counties typically have larger samples, which reduces uncertainty (and thus SE). Conversely, smaller counties have fewer respondents, making the estimate more variable and increasing SE.
-   What does this *overall* trend tell you about the relative ease of studying small vs large counties?
    -   It’s generally harder to study smaller counties in a statistically precise way because their estimates have higher uncertainty (larger SE). Larger counties yield more precise estimates (lower SE), making it easier to detect differences or changes in income.

# Going Further

<!-- ----------------------------------------------------------------------- -->

Now it's your turn! You have income data for every county in the United States: Pose your own question and try to answer it with the data.

### **q8** Pose your own question about the data. Create a visualization (or table) here, and document your observations.

```{r q8-task}
## "How do median household income estimates vary by family size across all U.S. counties?"
# Create a boxplot of income estimates by family size category
df_data |>
  ggplot(aes(x = category, y = income_estimate)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  labs(
    title = "Median Household Income by Family Size Category",
    x = "Family Size Category",
    y = "Median Household Income"
  ) +
  theme_minimal()
```

**Observations**:

-   As family size increases (from 2-person up to 6-person families), the median household income tends to rise
    -   larger families have higher total incomes, possibly because there are multiple earners or greater overall resources.
-   The interquartile range (IQR) and overall spread of incomes grow for larger families. This indicates greater variability in total income among families with more members.
-   While the medians shift upward, the boxplots still overlap. That means some 2-person families have higher incomes than some 5- or 6-person families, and vice versa.
-   The pattern could stem from having more wage earners in larger households, or differences in cost-of-living areas (e.g., urban vs. rural). Smaller families might also include retirees or single earners, which could lower their median incomes.

# References

<!-- ----------------------------------------------------------------------- -->

[1] Spielman SE, Folch DC, Nagle NN (2014) Patterns and causes of uncertainty in the American Community Survey. Applied Geography 46: 147–157. <pmid:25404783> [link](https://www.sciencedirect.com/science/article/pii/S0143622813002518?casa_token=VddzQ1-spHMAAAAA:FTq92LXgiPVloJUVjnHs8Ma1HwvPigisAYtzfqaGbbRRwoknNqZ6Y2IzszmGgIGH4JAPzQN0)
