---
title: "The Islands, Part 2: Study"
author: "Isabel de Luis, Sam Mendelson, Chris Nie, Jacob Prisament, Zhi Hong"
date: 202X-XX-XX
output:
  github_document:
    toc: true
prerequisites:
  - c10-islands-pt1
---

# Notes for Sam

-   Please fill in the "data scraping" section
-   Please add in your cleaning/pivot code!
-   Please do analysis section (q4)

*Purpose*: This is part 2 of 2. In part 1 you *planed* your statistical project, particularly your data collection. In this part you will give updates on your plan, and report your findings.

This challenge is deliberately shorter so you have time to collect and analyze your data.

*Important note*: While we expect that you did your data collection with your team, you need to complete your own individual report for c10.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

# Setup

<!-- ----------------------------------------------------------------------- -->

```{r setup}
library(tidyverse)
library(rsample)
```

### **q1** Planning a study (TEAMWORK)

While you provided this plan in c08 (Part 1), please include your plan here. In particular, describe how you updated your plan in response to feedback.

#### Population

For our study, we are going to study the islanders from every city in Ironbard. The six cities in Ironbard are:

-   Hofn
-   Vardo
-   Helvig
-   Bjurholm
-   Blondous
-   Helluland

#### Quantity of interest

We are interested in seeing how Net worth influences household size and the number of children an islander has.

#### Covariates

The covariates of our study are:

-   Net Worth
-   Socioeconomic factors
-   Occupation
-   Town

#### Observation or experiment?

The Islands allows you to ask islanders to complete tasks. If you just take measurements on your participants, then it's an *observational study*. But if you also introduce something that's meant to change the outcome of a measurement (e.g., drinking coffee before taking a test), that's called an *experimental study*. You need to decide whether your study is observational or experimental.

Since this study aims to investigate the relationship between an islander's net worth, the number of children they have, and the household size, it is an observational study since we can analyze the existing data without manipulating any variables.

#### Question / Hypothesis

Our question is how does an Ironbard's islander's net worth influence the number of children they have and the size of their household.

#### Sampling plan

We will be collecting data from a random sample of 20 households from each city in Ironbard. Since we are taking a random sample, the data will be representative of the general population, because random sampling ensures that each element in the population has an equal probability of being chosen and reduces the biases involved.

We chose to take samples from 20 households from each city (120 households total) due to the practical constraints of working within a simulated environment. A sample size of 20 is manageable for detailed analysis and will allow for good observation while maintaining feasiblity within the scop of the study.

To increase efficiency, we will use a Python script that interacts with the islands and quickly collects data.

##### Data Scraping (Sam please fill out)

### **q2** EDA

After scraping the data, we had six different CSV files with the data for each city. In order to use this data though, we needed to combine and pivot the data.

```{r load data}
# load from csvs
bjurholm_df <-read_csv("data/islands/bjurholm_out.csv")
blondous_df <-read_csv("data/islands/blonduos_out.csv")
helluland_df <- read_csv("data/islands/helluland_out.csv")
helvig_df <- read_csv("data/islands/helvig_out.csv")
hofn_df <- read_csv("data/islands/hofn_out.csv")
vardo_df <- read_csv("data/islands/vardo_out.csv")
```

```{r clean data}
# Sam please fill this in

# Replace the net worth value for anyone with 'NA' to 0
df <-
  df |> 
  mutate(NetWorth = replace_na(NetWorth, 0))

# People who are 1 have NA in their age value -> bandaid fix
df <-
  df |> 
  mutate(Age = replace_na(Age, 1))
```

## Sam please add pivoting explanation

## this chunk can be deleted once cleaning/pivot data section is updated

```{r load data}
load("data/final.RData")
```

Now that our data is cleaned, we want to take a look at it.

```{r summarize}
glimpse(df)

head(df)
```

From a glimpse, we see that our dataframe columns are:

-   Town
-   HouseID
-   HouseSize
-   Name
-   Net Worth
-   Occupation
-   Age
-   NumChildren
-   NumHouseOccupants

```{r summary statistics}
df |>
  summarise(mean_networth = mean(NetWorth),
            mean_age = mean(Age),
            mean_numChildren = mean(NumChildren),
            mean_numHouseOccupants = mean(NumHouseOccupants))

```

If we look at the mean of each variable of interest, we see that overall, islanders have around 1-2 children, there are 3-4 people in their household, and have a net worth of around \$2449.245.

```{r}
df |>
  ggplot(aes(x = NumChildren, color = Town)) +
  geom_density()
```

Looking at each town individually, in general for every town, the densities are right-skewed, suggesting that people have a lower number of children, with a few larger families in each town. Additionally, the curves overlap quite a bit, suggesting that the distribution for each town is somewhat similar. One thing to note about this plot is that it's looking at all the data we collected, which includes children. We can get a more accurate distribution of the number of children an islander has by filtering out every under the age of 18.

```{r}
df |>
  filter(Age >= 18) |>
  ggplot(aes(x = NumChildren, color = Town)) +
  geom_density()
```

When we take out the children from our density plot, we get a very different shape. Some towns appear to have multiple peaks, suggesting a multimodal distribution, which could indicate subgroups of families within a town. While many towns do cluster around 1-2 children, there is meaningful variation in distribution shape. However, the density plot may be showing multiple smaller peaks that may not be meaningful due to the smaller sampling size.

```{r}
df |>
  ggplot(aes(x = NetWorth, color = Town)) +
  geom_density()

df |>
  filter(Age >= 18) |>
  ggplot(aes(x = NetWorth, color = Town)) +
  geom_density()

```

Continuing with our densities analysis, we can look at the spread of the distribution of wealth on each island. All curves show a strong right skew, meaning most values cluster toward the lower end of net worth. Some towns have a noticeably broader spread in net worth (Blonduos, Vardo, Helvig), potentially indicating a more diverse population in terms of financial status.

If we look only at the adults in the dataset, we see a very similar plot.
```{r}
df |>
  ggplot(aes(x = NumHouseOccupants, color = Town)) +
  geom_density()
```
The next distribution we're going to look at is the number of household occupants. Several towns, like helluland and bjurholm appear to have more than one distinct peak. Helvig seems to have larger household size on average, while the other towns seem to have an average household size of 2-4.

```{r}
df |>
  ggplot(aes(x = Age, color = Town)) +
  geom_density()
```
The final distribution we're going to look at is age. The age range in the data is from 0 to around 80. All of the curves show a peak around young adult age (20 - 30) and a peak around middle age (40 - 50) or older adulthood (60+). This indicates that there is a disproportionately young population -- perhaps there was a baby boom in Ironbard about 20 years ago. 

The younger people in our datast may bias our observations -- since they are younger they have had less time to acquire wealth or have more children. While we make our key analyses we have to keep that in mind.

### **q3** Key Analyses

Present the key analyses that support your questions / hypotheses. This could include summary statistics (e.g., a proportion of Islanders), grouped summary statistics (e.g., a proportion for group A, and for group B), or visualizations (e.g., a histogram). This section should be short, and every analysis should directly relate to q4.

```{r}
df |>
  ggplot(aes(x = NumChildren, y = NetWorth)) +
  geom_col() +
  facet_wrap("Town", scales = "free")

df |>
  ggplot(aes(x = NumChildren, y = NetWorth)) +
  geom_col()
```

```{r}

# want to bin ages
# then find the mean networth for each unique number of children
# then find the mean number of children
df <- df |>
  mutate(age_bin = cut(
    Age, 
    breaks = seq(floor(0), ceiling(90), by = 10),
    include.lowest = TRUE, 
    right = FALSE  
  ))

summary_df <- df %>%
  group_by(age_bin, NumChildren) %>%
  summarise(mean_networth = mean(NetWorth, na.rm = TRUE), 
            .groups = 'drop')
summary_df |>
  ggplot(aes(x = age_bin, y = mean_networth, fill = factor(NumChildren))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Mean Networth by Age Bin and Number of Children",
    x = "Age Bin",
    y = "Mean Networth",
    fill = "Number of Children"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

```{r}
df |> 
  ggplot(aes(x = NumHouseOccupants, y = NetWorth)) +
  geom_col()
  
```

```{r}
df |>
  ggplot(aes(x = NumChildren, y = NumHouseOccupants, fill = NumHouseOccupants)) +
  geom_col(position = "dodge")

```
```{r}
df |>
  ggplot(aes(x = NumChildren, y = NumHouseOccupants, fill = NetWorth)) +
  geom_tile()
```

### **q4** Answers

Connect your Key Analyses to your questions / hypotheses. What did you conclude from your study?
