---
title: "Challenge 13: NYC Taxi Data"
author: "Isa de Luis & Sam Mendelson"
date: 2025-04-28
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if (!require(arrow)) install.packages("arrow")

library(tidyverse)
library(arrow)
library(ggplot2)
library(tidyr)
library(dplyr)
library(sf)

```

# Overview

This dataset, [from the NYC Taxi and Limousine Commission](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page), contains information about taxi trips in New York City.

New York City enacted a **congestion pricing** scheme in January 2025, which charges a fee for all vehicles entering Manhattan below 60th Street. While personal vehicles can get charged up to \$9/day (charged at first entry during a specified calendar day), congestion pricing for taxis works differently: they pay \$0.75/entry, which is then passed on to the passenger.

## Columns

| Field Name | Category | Description |
|--------------------------------|---------------------|-------------------|
| `VendorID` | Vendor Info | A code indicating the TPEP provider that provided the record. 1 = Creative Mobile Technologies, LLC 2 = Curb Mobility, LLC 6 = Myle Technologies Inc 7 = Helix |
| `tpep_pickup_datetime` | Time | The date and time when the meter was engaged. |
| `tpep_dropoff_datetime` | Time | The date and time when the meter was disengaged. |
| `passenger_count` | Passenger Info | The number of passengers in the vehicle. |
| `trip_distance` | Trip Info | The elapsed trip distance in miles reported by the taximeter. |
| `RatecodeID` | Fare Info | The final rate code in effect at the end of the trip. 1 = Standard rate 2 = JFK 3 = Newark 4 = Nassau or Westchester 5 = Negotiated fare 6 = Group ride 99 = Null/unknown |
| `store_and_fwd_flag` | Trip Info | Flag indicating whether the trip record was stored before being sent to the vendor.<br>Y = store and forward trip<br>N = not a store and forward trip |
| `PULocationID` | Location | TLC Taxi Zone where the taximeter was engaged. |
| `DOLocationID` | Location | TLC Taxi Zone where the taximeter was disengaged. |
| `payment_type` | Payment Info | Numeric code indicating how the passenger paid. 0 = Flex Fare 1 = Credit card 2 = Cash 3 = No charge 4 = Dispute 5 = Unknown 6 = Voided trip |
| `fare_amount` | Fare Info | Time-and-distance fare calculated by the meter. [More info](https://www.nyc.gov/site/tlc/passengers/taxi-fare.page) |
| `extra` | Fare Info | Miscellaneous extras and surcharges. |
| `mta_tax` | Tax Info | Tax automatically triggered based on the metered rate. |
| `tip_amount` | Fare Info | Tip amount – automatically populated for credit card trips. Cash tips are not included. |
| `tolls_amount` | Fare Info | Total amount of all tolls paid in trip. |
| `improvement_surcharge` | Fare Info | Improvement surcharge assessed at the flag drop since 2015. |
| `total_amount` | Fare Info | Total amount charged to passengers (excludes cash tips). |
| `congestion_surcharge` | Fare Info | Total amount collected for NYS congestion surcharge. |
| `airport_fee` | Fare Info | For pickups at LaGuardia and JFK airports only. |
| `cbd_congestion_fee` | Fare Info | Per-trip charge for MTA's Congestion Relief Zone (starting Jan. 5, 2025). |

## Congestion Pricing vs. Congestion Surcharge

You may have noticed that there are two different fees related to congestion in the dataset: `congestion_surcharge` and `cbd_congestion_fee`. The former is a fee that was introduced in 2019 to help fund the MTA, while the latter is a new fee that was introduced in 2025 as part of the congestion pricing scheme.

The politics of congestion pricing have been contentious in New York City for years. The plan was first proposed in 2007, but it wasn't until 2019 that the state legislature passed a bill allowing for congestion pricing. Tolls for private vehicles were sacked, however a \$2.50 per ride fee for taxi and ride share trips that originate in, end in, or pass through Manhattan below 96th street was established, which is in our dataset as the `congestion_surcharge`.

In 2025, the congestion pricing scheme was fully implemented, which is in our dataset as `cbd_congestion_fee`. This fee is charged to all vehicles entering Manhattan below 60th Street, regardless of whether they are taxis or private vehicles. The goal of the fee is to reduce traffic congestion and improve air quality in Manhattan.

For taxi trips, these two fees can add together. For example, a trip going from 112th Street to 16th Street would pay both fees, as the trip:

-   Begins above 96th street - no fee
-   Ends below 96th street, therefore the `congestion_surcharge` is applied
-   Needs to enter the congestion pricing zone (destination is below 60th street), therefore the `cbd_congestion_fee` is applied

This results in a total fee of \$3.25 (\$2.50 from the congestion surcharge and \$0.75 from the congestion pricing scheme) applied to the trip.

Read more about the congestion *surcharge* (2019) [here](https://www.tax.ny.gov/bus/cs/csidx.htm) and the congestion *pricing* (2025) [here](https://portal.311.nyc.gov/article/?kanumber=KA-03612).

## TLC Taxi Zones

The dataset tells us, roughly, where the trips start and end. New York City is broken up into TLC taxi zones, and we have the pickup and dropoff zone IDs in the dataset. The TLC taxi zones are not the same as the NYC neighborhoods, but they are similar. For example, the Upper West Side is broken up into several different TLC taxi zones, while the East Village is only one zone.

The Metropolitan Transportation Authority (MTA, New York City's public transportation authority) released a complimentary dataset that lists the zones included in the congestion pricing scheme and their corresponding geographic boundaries. This dataset is available [here](https://data.ny.gov/Transportation/MTA-Central-Business-District-Taxi-Zones/yfdc-w5jh/about_data).

The TLC also released a CSV containing each Zone's ID and the name/neighborhood name (it also includes whether green cabs can pick up there, but we're not using that data). We use this data to provide neighborhood names, which are more understandable than just the zone numbers. This data is available from the same site as the original dataset.

## Yellow and Green Cabs

When asked about the color of a New York City taxicab, the first color that likely comes to mind for many is yellow. When we see taxis in movies, TV, and, even just around NYC, they're almost always yellow.

However, NYC has two colors of cabs, with just one difference: yellow cabs can pick up passengers anywhere in NYC including airports, however green cabs (also knows as boro cabs, whose color is called "apple green") can not **pick up** passengers:

-   At NYC airports, including JFK Airport and LaGuardia Airport

-   In Manhattan, south of East 97th Street

-   In Manhattan, south of West 110th Street

Green cabs can, however, **drop off** passengers anywhere in NYC. For example, they could drive someone from Brooklyn to the Upper East Side, however not the other way around. Therefore, we decided to include green cab data in our dataset.

# Hypothesis

We hypothesize that, as a result of the congestion *pricing*, the number of trips that **terminate in the congestion *pricing* zones will decrease**.

If someone was, for example, going to take a taxi to Carnegie Hall (on 57th Street), the taxi driver might (smartly) recommend that they get out on 60th and just walk to Carnegie Hall to save the fee. In their self interest, they would want to get another trip sooner, and passengers may tip more knowing that the driver saved them money - if they tipped an extra \$0.75, then the entire fee would go to the driver.

# EDA

Here, we load the data (provided by the TLC (Taxi and Limousine Commission of New York State) as Apache Parquet files) into R.

Green

```{r}
df_feb_25_green <- read_parquet("data/project/green_tripdata_2025-02.parquet")
df_feb_25_yellow <- read_parquet("data/project/yellow_tripdata_2025-02.parquet")

df_feb_24_green <- read_parquet("data/project/green_tripdata_2024-02.parquet")
df_feb_24_yellow <- read_parquet("data/project/yellow_tripdata_2024-02.parquet")
```

Since each month and each taxi color is a separate file, we need to combine them. See [Yellow and Green Cabs] for more information on the difference between green and yellow cabs.

```{r}

df_feb_25_green <- 
  df_feb_25_green |>
  mutate(year = 2025,
         color = "green")

df_feb_24_green <- 
  df_feb_24_green |>
  mutate(year = 2024,
         color = "green")

df_feb_25_yellow <- 
  df_feb_25_yellow |>
  mutate(year = 2025,
         color = "yellow")

df_feb_24_yellow <- 
  df_feb_24_yellow |>
  mutate(year = 2024,
         color = "yellow")
df_combined <- bind_rows(df_feb_25_green, df_feb_25_yellow, df_feb_24_green, df_feb_24_yellow)
```

Let's take a quick look at the data we combined. We added two columns, `year` and `color` to differentiate the datasets. No `month` column is needed: both datasets are from February.

```{r}
glimpse(df_combined)
head(df_combined)
```

# Plots

## Dropoffs By Zone

```{r}
zone_lookup <- read_csv("data/project/taxi_zones.csv")
zone_lookup <- zone_lookup |>
  rename(LocationID = `Taxi Zone`)

zone_names <- read_csv("data/project/taxi_zone_lookup.csv")

zones_sf <- zone_lookup |>     
  st_as_sf(wkt = "Polygon", crs = 4326)        

df_zone_by_year <- df_combined |>
  group_by(year, DOLocationID) |>
  summarise(n_rides = n(), .groups = "drop")

df_zone_by_year <- df_zone_by_year |>
  group_by(year) |>
  mutate(
    frac  = n_rides / sum(n_rides),       
    pct   = frac * 100                    
  ) %>%
  ungroup() |>
  left_join(zone_names,
            by = c("DOLocationID" = "LocationID"))

map_data <- zones_sf |>
  left_join(df_zone_by_year,
            by = c("LocationID" = "DOLocationID"))

map_data |>
  ggplot() +
  geom_sf(aes(fill = pct), color = "grey70", size = 0.1) +
  scale_fill_viridis_c(
    option    = "magma",
    trans     = "sqrt",
    na.value  = "white"
  ) +
  facet_wrap(~ year) +
  labs(
    title    = "Feb Taxi Dropoffs in Manhattan by Zone",
    subtitle = "2024 vs 2025",
    fill     = "% of All Rides"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text    = element_blank(),
    axis.ticks   = element_blank(),
    panel.grid   = element_blank(),
    strip.text   = element_text(face = "bold")
  ) 

```

Here, we're looking at all the zones affected by congestion pricing, comparing February 2024 to February 2025. Congestion pricing has been enabled for the entirety of February 2025. The color indicates the percent of total trips that ended in each zone, with darker representing less and lighter representing more trips. The maps look very similar, with the most trips terminating in Zone 161 (Midtown Center).

Our central claim is that traffic to these zones would decrease after the congestion zone pricing. At a quick glance, this *might* be accurate, but let's verify further. The zones at the top of the congestion zone pricing are, in order: 50, 48, 163, 162, and 229.

```{r}
df_zone_by_year %>%
  filter(
    DOLocationID %in% c(50, 48, 163, 162, 229)
  ) %>%
  select(DOLocationID, Zone, year, pct, n_rides) %>%
  pivot_wider(names_from = year, values_from = c(pct, n_rides)) %>%
  mutate(
    pct_percent_change = (pct_2025 - pct_2024) / pct_2024 * 100,
    n_rides_percent_change = (n_rides_2025 - n_rides_2024) / n_rides_2024 * 100
  )
```

Looks like traffic in zones 48 and 50 (Clinton East and West) haven't changed much (\<1% each), however, in Midtown East and North along with Sutton Place/Turtle Bay North, traffic is down by significant percentages - 7.8%, 5.95%, and 6.8% respectively. While this does not establish a causal relationship between congestion pricing and this decrease in dropoffs, it is certainly interesting. (Note: we're defining traffic as the percentage of *total* *rides* that went to these zones.)

Let's take a look at the zones just *above* (north) of the three significantly affected zones (162, 163, and 229), which are zones 237, 141, and 140. (Technically zone 43, Central Park, is above zone 162, however this zone covers the entirety of Central Park, which is many times the sizes of all other zones.)

```{r}
df_zone_by_year %>%
  filter(
    DOLocationID %in% c(237, 141, 140)
  ) %>%
  select(DOLocationID, Zone, year, pct, n_rides) %>%
  pivot_wider(names_from = year, values_from = c(pct, n_rides)) %>%
  mutate(
    pct_percent_change = (pct_2025 - pct_2024) / pct_2024 * 100,
    n_rides_percent_change = (n_rides_2025 - n_rides_2024) / n_rides_2024 * 100
  )
```

Zone 237 is above a small portion of 163 and 162, and zones 141 and 140 are above zone 229 in the congestion pricing zone. Looks like traffic is down significantly in those three zones, too, providing evidence contrary to our central claim. However, these regions are fairly tall (\~10 streets) and, therefore, this evidence is far from conclusive.

While traffic is down, however, the total number of rides has increased. Interestingly, the traffic in zones 141 and 140 are up by 8.3% and 14% respectively, and their respective congestion pricing zone (229) only had growth of 10%. It is therefore theoretically possible that some of this extra growth in zone 140 over zone 229 (CP) is due to the congestion pricing, but there is far from enough evidence to make that a finding.

## Mean Fares, Tips, and Distances by **Pickup Location**

```{r}
# Mean Fare, Mean distance, mean tip percent per pickup zone
df_zone_stats <- df_combined |>
  filter(fare_amount > 0) |>
  group_by(year, PULocationID)|>
  summarise(
    median_fare    = median(fare_amount,      na.rm = TRUE),
    median_tip_pct = median(tip_amount/fare_amount*100, na.rm = TRUE),
    median_dist    = median(trip_distance,    na.rm = TRUE),
    .groups = "drop"
  )

zone_stats_sf <- zones_sf %>%
  left_join(df_zone_stats, by = c("LocationID" = "PULocationID"))

plot_map <- function(col, legend.title) {
  zone_stats_sf |>
  ggplot() +
    geom_sf(aes_string(fill = col), color = "grey80", size = 0.1) +
    facet_wrap(~year) +
    scale_fill_viridis_c(
      option   = "plasma",
      trans    = "sqrt",
      na.value = "white"
    ) +
    labs(fill   = legend.title,
         title  = legend.title,
         subtitle = "Feb 2024 vs Feb 2025") +
    theme_minimal() +
    theme(
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold")
    )
}

plot_map("median_fare",    "Median Fare ($)")
plot_map("median_tip_pct", "Median Tip (%)")
plot_map("median_dist",    "Median Distance (mi)")
```

We've also plotted the mean fares, tips, and distances by both pickup location and dropoff location. Here we're looking at the data by pickup location, and here are some observations:

-   In general, median fares decreased. This could indicate that New Yorkers are taking shorter trips to account for the fees. It is also possible that they are choosing taxis over Uber and Lyft, who pay double the congestion pricing surcharge (\$1.50 instead of \$0.75). For a \$10 trip, say, those 75 cents could account for a 7.5% increase in the fare. This would lower the median fare as more short trips are being taken in taxis.

-   However, median distance is around the same in many zones, conflicting with the above theory.

-   The median tip percentage also decreased in many zones. The largest decrease, by far, is in the Lower East Side, which is the lowest income area on our map ([source](https://bestneighborhood.org/household-income-new-york-ny/)), which might indicate new financial hardship for the city's least fortunate.

    -   Keep in mind that this metric only includes credit card tips. It is possible that people in the Lower East Side are simply choosing to tip in cash.

## Mean Fares, Tips, and Distances by **Dropoff Location**

```{r}
# Mean Fare, Mean distance, mean tip percent per dropoff zone
df_zone_stats <- df_combined |>
  filter(fare_amount > 0) |>
  group_by(year, DOLocationID)|>
  summarise(
    median_fare    = median(fare_amount,      na.rm = TRUE),
    median_tip_pct = median(tip_amount/fare_amount*100, na.rm = TRUE),
    median_dist    = median(trip_distance,    na.rm = TRUE),
    .groups = "drop"
  )

zone_stats_sf <- zones_sf %>%
  left_join(df_zone_stats, by = c("LocationID" = "DOLocationID"))

plot_map <- function(col, legend.title) {
  zone_stats_sf |>
  ggplot() +
    geom_sf(aes_string(fill = col), color = "grey80", size = 0.1) +
    facet_wrap(~year) +
    scale_fill_viridis_c(
      option   = "plasma",
      trans    = "sqrt",
      na.value = "white"
    ) +
    labs(fill   = legend.title,
         title  = legend.title,
         subtitle = "Feb 2024 vs Feb 2025") +
    theme_minimal() +
    theme(
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold")
    )
}

plot_map("median_fare",    "Median Fare ($)")
plot_map("median_tip_pct", "Median Tip (%)")
plot_map("median_dist",    "Median Distance (mi)")
```

These plots are the inverse idea of the ones above: they are based on the **dropoff** location, rather than the pickup location. Some observations from this data:

-   Median fares appear to have risen across the map. This is not a surprise at all: trips into this zone incur the \$0.75 fee, therefore that would raise fares in every zone.

    -   Since this shows data by dropoff location, it is possible for this map to show an increase in fares as trips come from outside the zone into the zone, however the pickup location map may not change as the fee is charged for *entry* to the congestion pricing zone.

-   Just as above, median tips decreased, with the largest change in the Lower East Side.

-   Median distance appears to be around the same for both 2024 and 2025.

# Conclusion

While it is possible that congestion pricing has had an effect on the number of trips into the new congestion pricing zone, from the data we saw, it is not attributable exclusively or in large part to the congestion pricing. We controlled for general traffic patterns by using the same month one year apart, but many things change in a year, and it is likely that other factors are at play causing the change we saw in traffic rates around the congestion pricing zone.
